---
title: 'Fundamentals of ggplot2'
output: html_document
aliases: /plotting-with-ggplot-the-basics/
---

# PART I - Creating your plot

```{r, include = FALSE}
library(ggplot2)
```

This lesson has been mainly borrowed from https://uc-r.github.io/ggplot_intro

## Dataset

```{r}
# View a summary of the data 

```


## Basics
In order to create a plot, you:

- Call the `ggplot()` function on your data which creates a blank canvas
- Specify aesthetic mappings, which control how you want to map variables to visual aspects
- Add new layers that are geometric objects which will show up on the plot

```{r}
# create canvas

# map variables of interest

# add geom object

```

When we add the geom layer we use the addition (+) operator. As you add new layers you will always use + to add onto your visualization.

## Aesthetic mappings

The aesthetic mappings take properties of the data and use them to influence visual characteristics, such as position, colour, size, shape, or transparency. Each visual characteristic can thus encode an aspect of the data and be used to convey information.

All aesthetics for a plot are specified in the `aes()` function call (later in this tutorial you will see that each geom layer can have its own aes specification).  

```{r}
# add a mapping from the class of the cars to a colour characteristic

```


Note that 'aesthetics' in ggplot refers to WHAT is plotted, now HOW it is plotted, unlike the word's usual meaning. Using the `aes()` function will cause the visual to be based on the *data* specified in the argument. For example, using `aes(colour = "blue")` won’t cause the geometry's colour to be 'blue', but will instead cause the visual to be mapped from the vector `c("blue")` — as if we only had a single type of engine that happened to be called 'blue'. If you wish to apply an aesthetic property to an entire geometry, you can set that property as an argument to the geom method, OUTSIDE of the `aes()` call:

```{r}
# illustrate the common mistake of trying to specify a colour within the aes() function

# here is the correct way

```

### Exercise:

1. Make a scatterplot of the relationship between the city and highway mileage of each car. Colour all the points red.

```{r}
# YOUR CODE HERE

```

2. Now, colour the points in your scatterplot by the number of cylinders each car has.

```{r}
# YOUR CODE HERE

```

## Specifying geometries

Building on these basics, ggplot2 can be used to build almost any kind of plot you may want. These plots are declared using functions.

The most obvious distinction between plots is what geometric objects (geoms) they include. ggplot2 supports a number of different types of geoms, including:

- `geom_point()`
- `geom_line()`
- `geom_smooth()`
- `geom_bar()`
- `geom_boxplot()`
- `geom_histogram()`
- `geom_polygon()`
- `geom_map()`

Each of these geometries will leverage the aesthetic mappings supplied. For example, you can map data to the location, colour, and shape of a geom_point (e.g., points can be circles or squares), or you can map data to the linetype of a geom_line (e.g., solid or dotted).

Most geoms require an x and y mapping as a bare minimum.

```{r}
# x and y mapping needed for geom_point and geom_smooth


# no y mapping needed for geom_bar and geom_histogram

```


### Exercise: 

1. Create a box-and-whisker plot of the engine displacement of the different drive types (front, rear, or 4wd).

```{r}
# YOUR CODE HERE

```

2. Create a box-and-whisker plot of the engine displacement of the different engine types (cylinders). What happens if your x variable is numeric? How might we fix that?

```{r}
# YOUR CODE HERE

```


## Layering geometries

What makes this approach really powerful is that you can add multiple geometries to a plot, allowing you to create complex graphics showing multiple aspects of your data.

```{r}
# plot with both points and smoothed line

```


Of course, the aesthetics for each geom can be different, so you could show multiple lines on the same plot (or with different colours, styles, etc). It’s also possible to give each geom a different data argument, so that you can show multiple data sets in the same plot.

For example, we can plot both points and a smoothed line for the same x and y variable but specify unique colours within each geom:

```{r}
# same as above, but points red and line blue

```

So if we specify an aesthetic within ggplot it will be passed on to each geom that follows. Or we can specify certain aes within each geom, which allows us to only show certain characteristics for that specific layer (i.e. geom_point).

```{r message=FALSE, warning=FALSE}
# colour aesthetic passed to each geom layer


# colour aesthetic specified for only the geom_point layer

```

### Exercise:

1. Take your boxplot from the previous exercise and add a layer that displays the values of the individual points on each box-and-whisker plot.

```{r}
#YOUR CODE HERE

```


## Position adjustments

In addition to a default statistical transformation, each geom also has a default position adjustment which specifies how different components should be positioned relative to each other. This position is noticeable in a geom_bar if you map a different variable to the colour visual characteristic:

```{r}
# bar chart of class, coloured by drive (front, rear, 4-wheel)

```

The geom_bar by default uses a position adjustment of 'stack', which makes each rectangle’s height proportional to its value and stacks them on top of each other. We can use the position argument to specify what position adjustment rules to follow:

```{r}
# position = "dodge": values next to each other


# position = "fill": percentage chart

```

Check the documentation for each particular geom to learn more about its positioning adjustments.

### Exercise:

1. Take your box-and-whisker plot from the previous exercise and use `position = "jitter"` in the point geometry layer. What happens?

```{r}

# YOUR CODE HERE


```

2. Create a histogram of the distribution of highway mileage values in mpg. Colour the bars by the drive type. Experiment with the position argument in the histogram geometry. What happens?

```{r}
# YOUR CODE HERE

```

# PART II - Refining your plot

## Managing scales

Whenever you specify an aesthetic mapping, `ggplot` uses a particular scale to determine the range of values that the data should map to. Thus when you specify

```{r}
# colour the data by engine type

```

ggplot automatically adds a scale for each mapping to the plot:

```{r}
# same as above, with explicit scales

```

Each scale can be represented by a function with the following name: `scale_`, followed by the name of the aesthetic property, followed by an _ and the name of the scale. A `continuous` scale will handle things like numeric data (where there is a continuous set of numbers), whereas a `discrete` scale will handle things like colours (since there is a small list of distinct colours).

While the default scales will work fine, it is possible to explicitly add different scales to replace the defaults. For example, you can use a scale to change the direction of an axis:

```{r}
# milage relationship, ordered in reverse

```

Similarly, you can use `scale_x_log10()` and `scale_x_sqrt()` to transform your scale. You can also use scales to format your axes:

```{r}

```

A common parameter to change is which set of colours to use in a plot. While you can use the default colouring, a more common option is to leverage the pre-defined palettes from [colourbrewer.org](https://colourbrewer2.org/). These colour sets have been carefully designed to look good and to be viewable to people with certain forms of colour blindness. We can leverage colour brewer palletes by specifying the `scale_colour_brewer()` function, passing the pallete as an argument.

```{r}
# default colour brewer


# specifying colour palette

```

Note that you can get the palette name from the colourbrewer website by looking at the scheme query parameter in the URL. Or see the diagram here and hover the mouse over each palette for the name.

You can also specify continuous colour values by using a gradient scale, or manually specify the colours you want to use as a named vector.

### Exercises

1. Produce a scatter plot with `displ` on the x axis and `hwy` on the y axis, and colour by `drv`. Next, go to [colourbrewer.org](https://colourbrewer2.org/) and find an appropriate colour palette that is colourblind safe. Update your plot with that pallete.

```{r}
# YOUR CODE HERE

```

2. Your plot should currently have three breaks on the y axis at 20, 30, and 40. Using `scale_y_continuous()`, try increasing the number of breaks so that there are breaks at 15, 20, 25, 30, 35 and so on. Hint: Check the help file using `?scale_y_continuous`.

```{r}
# YOUR CODE HERE

```

## Facets

Facets are ways of grouping a data plot into multiple different pieces (subplots). This allows you to view a separate plot for each value in a categorical variable. You can construct a plot with multiple facets by using the `facet_wrap()` function. This will produce a “row” of subplots, one for each categorical variable (the number of rows can be specified with an additional argument):

```{r}

```

You can also `facet_grid()` to facet your data by more than one categorical variable. Note that we use a tilde (`~`) in our `facet` functions. With `facet_grid()` the variable to the left of the tilde will be represented in the rows and the variable to the right will be represented across the columns.

```{r}

```

### Exercises

1. Produce a scatter plot with a smoothed line with `displ` on the x axis and `hwy` on the y axis. Facet the plot by `drv` using `facet_grid()`.

```{r}
# YOUR CODE HERE

```

2. Now try setting `scales='free'` inside `facet_grid()`? What has changed?

```{r}
# YOUR CODE HERE

```

## Labels & annotations

Textual labels and annotations (on the plot, axes, geometry, and legend) are an important part of making a plot understandable and communicating information. Although not an explicit part of the Grammar of Graphics (the would be considered a form of geometry), `ggplot` makes it easy to add such annotations.

You can add titles and axis labels to a chart using the `labs()` function (not `labels`, which is a different R function!):

```{r}

```
